Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2024-12-30 01-introduction.qmd
uodates para respuestas reviewers

[[no sé si de pronto aclarar más la idea de que no es posible aislar el efecto COVID propiamente tal ...]]

>>>>

Against this backdrop, Chile serves as an example among Latin American countries due to its strong sanitary measures. These measures significantly impacted the labor market, leading to important consequences for unemployment and income distribution. It is importantto approach any claims about the influence of the COVID-19 pandemic with caution. Our findings are situated within a broader context of increasing socioeconomic risks and vulnerabilities, which were heightened by the sanitary restrictions implemented during the pandemic.

### 2024-12-30 02-antecedentes.qmd
uodates para respuestas reviewers

@@

The present research aims at identifying changes in the justification of educational inequalities in Chile, with a particular focus in its relationship with meritocratic perceptions. The time frame for this analysis covers between 2016 and 2023, by using longitudinal panel survey data provided by the ELSOC database (Estudio Longitudinal Social de Chile). Although this period comprises the COVID pandemic, the changes in the variables analyzed are not possible to be attributed only to this event in terms of a health crisis, as several economic, social and cultural processes overlapped worldwide. Besides, in the specific case of Chile there were two constitutional processes happening in this very same period after a massive social upheaval. [[redondear-idea]]

As there are measures before and after the pandemic period, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021]. Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period. Consistently, our first hypothesis is:

>>>>

Empirical cross-sectional evidence from Chile reveals two significant trends: low public support for the perception of meritocracy and limited justification for unequal access to social services such as education and health. However, the relationship between meritocracy and the justification of inequality within the educational domain remains underexplored. For instance, @castillo_clivajes_2013 analyzed data from the International Social Survey Programme (ISSP) in 1999 and 2009, highlighting consistently low support for inequality based on resource allocation or individual merit. Similarly, @Castilloetal2019 demonstrated widespread skepticism among Chileans regarding the meritocratic system, with many disagreeing with the notion that rewards are fairly distributed based on talent and effort.

@@

The present research aims at identifying changes in the justification of educational inequalities in Chile, with a particular focus in its relationship with meritocratic perceptions. The time frame for this analysis covers between 2016 and 2023, by using longitudinal panel survey data provided by the ELSOC database (Estudio Longitudinal Social de Chile). Although this period comprises the COVID pandemic, the changes in the variables analyzed are not possible to be attributed only to this event in terms of a health crisis, as several economic, social and cultural processes overlapped worldwide. Besides, in the specific case of Chile there were two constitutional processes happening in this very same period after a massive social upheaval.

As there are measures before and after the pandemic period, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021].

Within this framework, we point out that periods of rising socioeconomic and health vulnerabilities could weaken the justification of social inequalities. This change would be at least in part linked to a challenge of the meritocratic ideals. Chile provides a significant case for studying inequality, as it exemplifies a prosperous yet highly unequal society within Latin America. Under such conditions, a shock like the COVID-19 pandemic posed a considerable threat to economically vulnerable groups and those in precarious labor market positions. Chile implemented sanitary controls early in the pandemic, initiating lockdowns on March 18, 2020, and lifting all restrictions by March 5, 2023. In this context, Chile stands out as one of the strictest countries globally and within South America in terms of pandemic-related policies. According to the Stringency Index [@hale_global_2021]—ranging from 0 to 100 (100 being the strictest) and encompassing measures like school closures, workplace shutdowns, and travel bans—the global median was 42.54, while Chile scored 58.1. This placed Chile 36th among 184 countries and 4th in South America, following Peru (60.1), Suriname (60.5), and Venezuela (80.5).

Within the period analyzed (2016-2022), an important cycle of protests known as the "social outburst" took place between October 2019 and March 2020 [@somma_no_2021]. However, the protest drastically decreased as a result of the lockdown [@joignant_informe_2020]. In terms of public views on economic inequality, it has been documented that critical views toward inequality increased in 2019 and deepened under the pandemic [@coes_radiograficambio_2023], which we argue can be in part explained by the increased risk exposure and vulnerability in the context of the COVID-19 pandemic. 
Hereby, a relevant aspect is the transformation of the labor market and its impact on income distribution [@ministerio_de_desarrollo_social_y_familia_resultados_2022;@instituto_nacional_de_estadisticas_resultados_2022,]. In 2015, the poverty rate was 11.2%, decreasing to 8.5% by 2017, with the 10/10 ratio at 31.6. In 2020, during the COVID-19 pandemic, poverty increased to 10.7%, and the 10/10 ratio surged to 301.7, while unemployment reached 10.5%, the highest level in a decade. By 2022, poverty declined to 6.5%, the 10/10 ratio improved to 61.2, and unemployment decreased to 7.9%, rising slightly to 8.7% in 2023.

Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period. Consistently, our first hypothesis is:

### 2024-12-30 03-methods.qmd
uodates para respuestas reviewers

For testing the hypothesis regarding status we included educational level and household income quintiles, as well as subjective social status. Finally, political identification on a left-right scale, age groups, and gender were included as control variables. @tbl-descriptives shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.

>>>>

For testing the hypothesis regarding status we included educational level and household income quintiles. Also, we included subjective social status as it has been argued that perceived social status is a relevant predictor of attitudes toward economic inequality [@schneider_poverty_2015;@castillo_meritocracia_2019]. In addition, subjective measures can complement objective measures in predicting life chances [@oesch_subjective_2023], which are arguably connected to how individuals experience economic inequality. In this regard, it has been argued that how people form their views on their standing in society results from experiences with direct socioeconomic circumstances and social comparison processes with reference groups [@Condon2020]. Finally, political identification on a left-right scale, age groups, and gender were included as control variables. @tbl-descriptives shows the whole set of independent variables, the corresponding items,  response categories, and frequencies.
